\large Neural and Symbolic Optimization of Cosmological Particle-Mesh Simulation \vspace{0.3em}

### 2024-04-25 docs/papers/sto/so.tex
Update paper draft

\author[0000-0002-9300-2632]{\normalsize Yucheng Zhang}
\affiliation{Peng Cheng Laboratory, Shenzhen, Guangdong 518066, China}
\affiliation{Center for Cosmology and Particle Physics, New York
University, New York, New York 10003, USA}

@@

@@

% Main text: up to 3500 words: introduction + results + discussion
\vspace{1em}
\section{Introduction}
% without heading

% AI for science and cosmology
Rapid advances in artificial intelligence (AI) and more specifically
deep learning have provided powerful tools that are revolutionizing
scientific research across disciplines.
AI for science benefits from not only the state-of-art architectures of
deep neural networks (NNs), but also more fundamental computational
tools and hardwares like automatic differentiation (AD) and GPUs.
One of the very promising areas of AI in Cosmology is the numerical
simulation of the Universe.

@@

Cosmological $N$-body simulations evolve dark matter particles under the
gravitational interaction.
The dynamics of the dark matter is governed by the Poisson equation
%
\begin{equation}
  \nabla^2 \varphi(\vx) = \frac32 \Omegam H_0^2 \delta(\vx),
\end{equation}
%
where $\Omegam$ is the matter fraction today, $H_0$ is the Hubble
constant, and the two fields $\varphi(\vx)$ and $\delta(\vx)$ are the
gravitational potential and matter density fluctuation respectively.
In Fourier space, the gravitational force field $\vF = - \vnabla
\varphi$ follows
%
\begin{equation}
\vF(\vk) = - i \vk \varphi(\vk) =
                  i \frac32 \Omegam H_0^2 \,\frac{\vk}{k^2} \delta(\vk),
\label{force}
\end{equation}
%
where the Laplacian operator reduces to simple multiplications.
The force field in configuration space $\vF(\vx)$ can then be recovered
with an inverse Fourier transform.
The particle-mesh (PM) method is a fundamental and efficient approach to
solving the Poisson equation given numerous particles.
In order to use the fast Fourier transform (FFT), PM estimates the
matter density field on a regular spatial grid by scattering the
particles onto an auxiliary mesh.
The cloud-in-cell (CIC), or trilinear, interpolation scheme
\citep{HockneyEastwood1988} has been widely used to compute the mass
fractions of a particle at $\vx'$ scattered to a nearby mesh point
at $\vx$
%
\begin{equation}
  W(\vx, \vx') = \prod_{i=1}^3
    \max \Bigl( 1 - \frac{|x_i-x'_i|}{l_c}, 0 \Bigr),
  \label{cic}
  \end{equation}
%
where $l_c$ is the mesh cell size.
Other scattering schemes include the nearest grid point (NGP) and
triangle shaped cloud (TSC) etc.
The gravitational forces on the mesh points are computed in the Fourier
space and transformed back to the configuration space, which are then
grathered to form the interpolated force for each particle, with the
same weights as in the scattering process.
Although being computationally efficient, the accuracy of the
gravitational force estimated for each particle depends on the
resolution of the mesh.
The PM method is mainly limited by its resolution on small scales, i.e.
scales around and below the mesh cell size.
This is due to the loss of information during the scattering and
gathering process between the continuously distributed particles and the
regular discrete mesh.

In this work, we sharpen the gravitational force by embedding NNs in the
force equation and train them for the spatial optimization (SO) of the
PM simulation.
NNs are flexible and efficient universal function approximators
\citep{Hornik1989}, which can be used to learn physical processes
when embedded in numerical simulations.
NNs are typically trained with gradient based optimizers, which requires
the simulations to be differentiable.
The differentiable cosmological simulation with adjoint method
\citep{Li2022a} and the corresponding PM $N$-body library \pmwd\
\citep{Li2022b} lay the foundation of the gradient-based optimization
discussed in this work.
We further interpret the trained NNs using symbolic regression (SR).
Optimizing the accuracy of fast and differentiable simulations is
crucial for their applications in, e.g. generation of mocks for
cosmological surveys, simulation-based inference or field-level
inference.

@@

To improve the accuracy of the PM solver, we introduce SO force
sharpening by modifying the gravitational force equation \eqref{force}
to
%
\begin{equation}
F(\vk)_i = i \frac32 \Omegam H_0^2 \,\frac{k_i}{k^2}\,
                    f(k_i; \vvartheta) g(\vk; \vvartheta) \delta(\vk),
\end{equation}
%
where $f$ and $g$ are nonlinear functions of the wavevector components
and all possible relevant features $\vvartheta$, which are approximated
with NNs.
$k_i$ is a component of $\vk$, with $i\in\{1,2,3\}$ denoting the 3
principal axes of the simulation box.

Gravity obeys the $\mathrm{O}(3)$ rotational symmetry, which is however
broken by the PM solver to the O\textsubscript{h} discrete symmetry with
48 elements involving permutations and reversals of the 3 principal axes
of the simulation box.
The PM evaluation of each force component can further break the symmetry
along that axis, leaving the remaining 2 permutationally symmetric.
Therefore, we implement $f$ and $g$ to be even functions of the
wavevector components, and $g$ to be permutation equivariant in all 3
components, so that $f g$ is the most general modification to the force
kernel.
An even and real kernel in configuration space is also even in Fourier
space.
We take the absolute value of the wavevectors for the evenness, and sort
the 3 wavevector compontents to achieve the permutation equivariance.

@@

Our target data samples are generated with the \GADGET\ simulation code
\citep{GADGET-4}.
We run 512 \GADGET\ simulations that cover a wide range of
configurations and cosmological parameters, as listed in
\autoref{tab:param}.
\YL{
Each simulation evolves $128^3$ particles from the initial time $a_\ic$
to the final time $a_\f=1$, saving 64 snapshots linearly spaced in scale
factor from $a_\ic + \gamma \Delta a$ to $1 - (1 - \gamma) \Delta a$.
We determine $a_\ic$ for each configuration depending on its cosmology
and mean particle spacing \citep{MichauxEtAl2021}, \YL{(YL elaborate why
and how here.)} which then sets the snapshot spacing $\Delta a = (1 -
a_\ic) / 64$.
To sample different alignment between the snapshots and the start/stop
times and avoid possible related numerical errors in \pmwd, we introduce
a fractional random offset $\gamma$ that varies between 0 and 1 for
different simulations.
}
The curvature $\OmegaK$ is related to the separate universe simulation
\citep{LiEtAl2014, WagnerEtAl2015}.
\YZ{We refer to this training dataset of 512 \GADGET\ simulations
sampled with Sobol sequence as \GSDATA.}
On average each \GADGET\ simulation takes roughly 2K CPU hours to run,
and the generation of the whole \GSDATA\ dataset consumes 1M CPU hours.

@@

\subsection{Spatial Optimization Neural Networks}

@@

For both the $f$ and $g$ spatial optimization NNs (SONNs), the input
features consist of tabular data, therefore we adopt multilayer
perceptron (MLP, i.e. NN with fully connected layers) structures to
learn the nonlinear functions.
Both $f$ and $g$ networks have $XXX$ hidden layers with $XXX\times n_{\rm
input}$ nodes for each layer, where $n_{\rm input}$ is the size of the
input layer, which is [...] and [...] for $f$ and $g$ respectively.
The rectified linear unit (ReLU) activation function is used for hidden
layers.
We further apply the exponential function to the output as a regulator
for positive output.
Being embedded in the differentiable simulation library \pmwd, the NNs
can be trained using gradient-based optimization methods as in usual
deep learning tasks.
The parameters of the NNs, which enter the force evaluation in every
time integration step, can be treated just the same as other
cosmological parameters.
The objective gradients are accumulated, based on which the parameters
are then optimized.

@@

\subsection{Symbolic Expressions}

@@

\subsection{Dark Matter Density Fields}

Improvements of SO on the dark matter density fields are the first
things to check.
We construct these fields by scattering the particles onto a mesh.
To compare two fields $\hat f$ and $f$, we introduce for each mode the
transfer function (TF), which is defined as the square root ratio of the
auto power spectrum:
%
\begin{equation}
T(\vk) \triangleq
\sqrt{\frac{|\hat f(\vk)|^2}{|f(\vk)|^2}},
\label{T}
\end{equation}
%
and the correlation coefficient (CC):
%
\begin{equation}
r(\vk) \triangleq
\frac{\Re \bigl[ \hat f(\vk) f^*(\vk) \bigr]}
     {\sqrt{|\hat f(\vk)|^2 |f(\vk)|^2}},
\label{r}
\end{equation}
%
where the numerator is the real part of the cross power spectrum.
The two fields are identical when both $T$ and $r$ are unity.
The PM method is accurate on scales much larger than the mesh cell size,
therefore we expect $T$ and $r$ to be very close to being unity both
before and after the SO with force sharpening.

In \autoref{fig:tfcc}, we show the results of TF and CC for both
training and test data samples.
To show different box sizes, we normalize the wavenumber by the Nyquist
wavenumber $k_{\rm Nyquist} = \pi / l_p$ of the particle grid, where
$l_p$ is the particle spacing.
With discrete Fourier transform (DFT) on a regular grid of length $N_c$
and cell size $l_c$, the range of sampled 1D discrete wavenumbers is
limited by the fundamental and Nyquist values, i.e. $k_{\rm fund} =
2\pi/(l_c N_c)$ and $k_{\rm Nyquist} = \pi/l_c$, where $l_c N_c$ is
simply the box size.
Then the normzalized wavenumber $k/k_{\rm Nyquist}$ gives the same range
$[2/N_c, 1]$ for different box sizes.
For the 3D wavenumber $k=\sqrt{\sum_{i=1}^{3}k_i^2}$, the maximum
normalized value is then $\sqrt{3}$.
Compared to original PM, the TF of PM-SONN is improved to be consistent
with \GADGET\ within 10\% up to the particle grid Nyquist wavenumber.

The effect of force sharpening can also be visually shown by looking
directly at the higher resolution of the density field
(\autoref{fig:denslab}) on small scales.
Compared to orginal PM, PM-SONN is more similar to \GADGET\ with sharper
big halos and more small-scale structures.

\subsection{Dark Matter Halos}

\subsection{Benchmarks}
We compare the computational cost of \GADGET, original PM, PM-SONN, and
PM-SOSR.
In the generation of our training data, on average each \GADGET\
simulation takes about 2K CPU hours.

@@

\vspace{1em}
\subsection{Configurations and Data Samples}

@@

\vspace{1em}
\textit{\large Data Availability:}
We make the \GSDATA\ simulation data set public at ...

>>>>

\author[0000-0002-9300-2632]{\normalsize Yucheng Zhang}
\affiliation{Peng Cheng Laboratory, Shenzhen, Guangdong 518000, China}
\affiliation{Center for Cosmology and Particle Physics, New York
University, New York, New York 10003, USA}

@@

% \author{\normalsize Heung-Yeung Shum}
% \affiliation{International Digital Economy Academy, Shenzhen, Guangdong
% 518000, China}
% \affiliation{Tsinghua University, Beijing 100084, China}
% \affiliation{The Hong Kong University of Science and Technology,
% Hong Kong 999077, China}

@@

% Main text: up to 3500 words: introduction + results + discussion
\vspace{1em}
\section{Introduction}
\label{sec:intro}
% without heading

% AI for science and cosmology
Rapid advances in artificial intelligence (AI) and more specifically
deep learning have provided powerful tools that are revolutionizing
scientific research across disciplines.
AI for science benefits from not only the state-of-art architectures of
deep neural networks (NNs), but also more fundamental computational
tools and hardwares like automatic differentiation (AD) and GPUs.
One of the very promising areas of AI in cosmology is the numerical
simulation of the Universe.

@@

Cosmological $N$-body simulations evolve dark matter particles under the
gravitational interaction, which at the field level is described by the
Poisson equation
%
\begin{equation}
  \nabla^2 \varphi(\vx) = \frac32 \Omegam H_0^2 \delta(\vx),
\end{equation}
%
where $\Omegam$ is the matter fraction today, $H_0$ is the Hubble
constant, and the two fields $\varphi(\vx)$ and $\delta(\vx)$ are the
gravitational potential and matter density fluctuation respectively.
In Fourier space, the gravitational force field $\vF = - \vnabla
\varphi$ follows
%
\begin{equation}
\vF(\vk) = i \frac32 \Omegam H_0^2 \,\frac{\vk}{k^2} \delta(\vk),
\label{force}
\end{equation}
%
where the Laplacian operator reduces to simple multiplications of the
wavevectors.
The force field in configuration space $\vF(\vx)$ can then be recovered
with an inverse Fourier transform of $\vF(\vk)$.
The particle-mesh (PM) method is a fundamental and efficient approach to
solving the Poisson equation in the $N$-body simulation.
In order to perform the transformation between configuration and Fourier
spaces efficiently using the fast Fourier transform (FFT), PM estimates
the matter density field on a regular grid by scattering the particles
onto an auxiliary mesh.
The cloud-in-cell (CIC), or trilinear, interpolation scheme
\citep{HockneyEastwood1988} has been widely used to compute the mass
fractions of a particle at $\vx'$ scattered to a nearby mesh point
at $\vx$
%
\begin{equation}
  W(\vx, \vx') = \prod_{i=1}^3
    \max \Bigl( 1 - \frac{|x_i-x'_i|}{l_c}, 0 \Bigr),
  \label{cic}
  \end{equation}
%
where $l_c$ is the mesh cell size.
Other scattering schemes include the nearest grid point (NGP) and
triangle shaped cloud (TSC) etc.
The gravitational force field is evaluated in the Fourier space and
transformed back to the configuration space on the mesh points, which
are then grathered to form the interpolated force for each particle,
with the same weights as in the scattering process.
Although being computationally efficient, the accuracy of the
gravitational force estimated for each particle depends on the
resolution of the mesh.
The PM method is mainly limited by its accuracy on small scales around
and below the mesh cell size, due to the loss of information during the
scattering and gathering process between the continuously distributed
particles and the regular discrete mesh.

In this work, we perform the spatial optimization (SO) of the PM
simulation by sharpening the gravitational force.
We embed NNs in the force equation \eqref{force} and train them towards
the targets that are more accurate on small scales.
NNs are flexible and efficient universal function approximators
\citep{Hornik1989}, which can be used to learn physical processes in
numerical simulations.
It is necessary for the simulations to be differentiable in order to
train the NNs with gradient based optimizers \citep[e.g.][]{Adam}.
We develop our training pipeline of the SO NNs in the framework of
the differentiable cosmological simulation with adjoint method
\citep{Li2022a} and the corresponding PM $N$-body library \pmwd\
\citep{Li2022b}.
To gain further physical insights into the PM method, we fit the trained
NNs using symbolic regression (SR).
Optimizing the accuracy of fast and differentiable simulations like
\pmwd\ is crucial for their applications in, e.g. generation of mocks
for cosmological surveys, simulation-based inference or field-level
inference.

@@

To improve the accuracy of the PM solver, we introduce SO force
sharpening by modifying the gravitational force equation \eqref{force}
to
%
\begin{equation}
F(\vk)_i = i \frac32 \Omegam H_0^2 \,\frac{k_i}{k^2}\,
                    f(k_i; \vvartheta) g(\vk; \vvartheta) \delta(\vk),
\end{equation}
%
where $f$ and $g$ are nonlinear functions of the wavevector components
and all possible relevant features $\vvartheta$, which are approximated
with NNs.
$k_i$ is a component of $\vk$, with $i\in\{1,2,3\}$ denoting the 3
principal axes of the simulation box.
Gravity obeys the $\mathrm{O}(3)$ rotational symmetry, which is however
broken by the PM solver to the O\textsubscript{h} discrete symmetry with
48 elements involving permutations and reversals of the 3 principal axes
of the simulation box.
The PM evaluation of each force component can further break the symmetry
along that axis, leaving the remaining 2 permutationally symmetric.
Therefore, we implement $f$ and $g$ to be even functions of the
wavevector components, and $g$ to be permutation equivariant in all 3
components, so that $f g$ is the most general modification to the force
kernel.
An even and real kernel in configuration space is also even in Fourier
space.
We take the absolute value of the wavevectors for the evenness, and sort
the 3 wavevector compontents to achieve the permutation equivariance.

For both the $f$ and $g$ spatial optimization NNs (SONNs), their input
features consist of tabular data, therefore we adopt multilayer
perceptrons (MLPs, i.e. NNs with fully connected layers) to learn the
nonlinear functions.
Both $f$ and $g$ networks have $XXX$ hidden layers with $XXX\times
n_{\rm input}$ nodes for each layer, where $n_{\rm input}$ is the size
of the input layer, which is [...] and [...] for $f$ and $g$
respectively.
The rectified linear unit (ReLU) activation function is used for hidden
layers.
We further apply the exponential function to the output as a regulator
for positive output.
Being embedded in the differentiable simulation library \pmwd, the NNs
can be trained using gradient-based optimization methods as in usual
deep learning tasks.
The parameters of the NNs, which enter the force evaluation in every
time integration step, can be treated just the same as other
cosmological parameters.
The objective gradients are accumulated, based on which the parameters
are then optimized.

@@

Our target data samples are generated with the \GADGET\ simulation code
\citep{GADGET-4}, configured with high force accuracy settings.
We run 512 \GADGET\ simulations that cover a wide range of
configurations and cosmological parameters, as listed in
\autoref{tab:param}.
\YL{
Each simulation evolves $128^3$ particles from the initial time $a_\ic$
to the final time $a_\f=1$, saving 64 snapshots linearly spaced in scale
factor from $a_\ic + \gamma \Delta a$ to $1 - (1 - \gamma) \Delta a$.
We determine $a_\ic$ for each configuration depending on its cosmology
and mean particle spacing \citep{MichauxEtAl2021}, \YL{(YL elaborate why
and how here.)} which then sets the snapshot spacing $\Delta a = (1 -
a_\ic) / 64$.
To sample different alignment between the snapshots and the start/stop
times and avoid possible related numerical errors in \pmwd, we introduce
a fractional random offset $\gamma$ that varies between 0 and 1 for
different simulations.
}
The curvature $\OmegaK$ is related to the separate universe simulation
\citep{LiEtAl2014, WagnerEtAl2015}.
We refer to this training dataset of 512 \GADGET\ simulations
sampled with Sobol sequence as \GSDATA.
On average each \GADGET\ simulation takes roughly 2K CPU hours to run,
and the generation of the whole \GSDATA\ dataset consumes 1M CPU hours.

@@

\subsection{Optimized Dark Matter Density Fields}

As discussed in \autoref{sec:intro}, our main motivation on performing
SO of the PM method is to improve its accuracy on simulating dark matter
distribution on small scales.

The effect of SO can be visually shown by looking directly at the
density field (\autoref{fig:denslab}) on small scales.
We construct the dark matter density fields by scattering the particles
onto another mesh.
Compared to orginal PM, optimized PM with SONN is more similar to
\GADGET\ with sharper big halos and more small-scale structures.

To compare two fields $\hat f$ and $f$, we introduce for each mode the
transfer function (TF), which is defined as the square root ratio of the
auto power spectrum:
%
\begin{equation}
T(\vk) \triangleq
\sqrt{\frac{|\hat f(\vk)|^2}{|f(\vk)|^2}},
\label{T}
\end{equation}
%
and the correlation coefficient (CC):
%
\begin{equation}
r(\vk) \triangleq
\frac{\Re \bigl[ \hat f(\vk) f^*(\vk) \bigr]}
     {\sqrt{|\hat f(\vk)|^2 |f(\vk)|^2}},
\label{r}
\end{equation}
%
where the numerator is the real part of the cross power spectrum.
The two fields are identical when both $T$ and $r$ are unity.

In \autoref{fig:tfcc}, we show the results of TF and CC for both
training and test data samples.
With discrete Fourier transform (DFT) on a regular grid of length $N_c$
and cell size $l_c$, the range of sampled 1D discrete wavenumbers is
limited by the fundamental and Nyquist values, i.e. $k_{\rm fund} =
2\pi/(l_c N_c)$ and $k_{\rm Nyquist} = \pi/l_c$, where $l_c N_c$ is
simply the box size.
To show different box sizes, we normalize the wavenumber by the Nyquist
value $k_{\rm Nyquist} = \pi / l_c$ of the force mesh grid, where
$l_c$ is the cell size.
Then the normzalized wavenumber $k/k_{\rm Nyquist}$ gives the same range
$[2/N_c, 1]$ for different box sizes.
For the 3D wavenumber $k=\sqrt{\sum_{i=1}^{3}k_i^2}$, the maximum
normalized value is then $\sqrt{3}$.
The original PM method is accurate on scales much larger than the mesh
cell size, therefore $T$ and $r$ are already very close to one without
the application of SONN.
Compared to original PM, the TF of PM-SONN is improved to be consistent
with \GADGET\ within 10\% up to the particle grid Nyquist wavenumber.

\subsection{Optimized Dark Matter Halo Distributions}

\subsection{Neural Networks and Symbolic Expressions}

@@

@@

@@

\subsection{Benchmarks}
We compare the computational cost of \GADGET, original PM, PM-SONN, and
PM-SOSR.
In the generation of our training data, on average each \GADGET\
simulation takes about 2,000 CPU hours.

@@

\vspace{1em}
\subsection{Training Data and Configurations}

@@

\vspace{1em}
\textit{\large Data Availability:}
We make the \GSDATA\ simulation dataset public at ...
